---
title: "ZhiLiu_HW5"
date: "October 1, 2016"
output: html_document
#  pdf_document:
#    toc: TRUE
#    toc_depth: 3
#    fig_caption: True

---

\newpage

## 6.3

a. See below:

```{r}
library(nnet)
data <-
array(c(23, 4, 2, 2, 8,
        7, 0, 1, 3, 5,
        5, 11, 1, 0, 3,
        13, 8, 6, 1, 0,
        5, 11, 2, 1, 5,
        8, 7, 6, 3, 5,
        16, 19, 1, 2, 3,
        17, 1, 0, 1, 3),
      dim = c(5, 2, 4),
      dimnames = list(
          PFC = c("F", "I", "R", "B", "O"),
          Length = c("S", "L"),
          Lake = c("H", "O", "T", "G")))
data.ftable <- ftable(data, row.vars = c("Lake", "Length"))
data.frame.w.freq <- data.frame(data.ftable)
data.frame.w.freq$Lake = relevel(data.frame.w.freq$Lake, ref = "G")
data.frame.w.freq$Length = relevel(data.frame.w.freq$Length, ref = "L")
data.frame.w.freq$PFC = relevel(data.frame.w.freq$PFC, ref = "O")
m <- multinom(PFC ~ Lake + Length, weights = Freq, data = data.frame.w.freq)
summary(m)
```

The prediction equations are:


$$\log \frac{P(PFC = F)}{P(PFC = O)} = 1.904343 - 0.8263115 LakeH - 0.005792737 LakeO - 1.516461 LakeT - 0.3315514 LengthS $$
$$\log \frac{P(PFC = I)}{P(PFC = O)} = 0.3552765 - 2.4845554 LakeH + 0.931497563 LakeO - 0.3944040 LakeT + 1.1266378 LengthS $$
$$\log \frac{P(PFC = R)}{P(PFC = O)} = -1.4101724 + 0.4165138 LakeH + 2.453110922 LakeO + 1.4188174 LakeT - 0.6829016 LengthS $$
$$\log \frac{P(PFC = B)}{P(PFC = O)} = -0.1888306 - 0.1310251 LakeH - 0.658687286 LakeO - 0.4285436 LakeT - 0.9622500 LengthS $$


b. See below:

```{r}
predicted.probs <- predict(m, type='probs')
# row number 2 and 6 correspond to the probability of PFC = F
# of (LakeO, LengthS) and (LakeO, LengthL)
predicted.probs[c(2, 6), 'F'] 
```

A larger length makes it more likely for alligators in Lake Oklawaha to choose fish as their primary food choice.

## 6.7

a. In a ordinal multicategory logit model, the $\beta$s for different response categories are the same. Here income is treated as an ordinal variable and gets assigned numerical values, so only one variable is enough. The number of intercepts is number of categories - 1 = 2.

b. The $\log Odds$ of happiness level $Y \le j\ (j = not, pretty)$ is 0.1117 unit less with every unit of income increase. However, the p-value is rather large. So the effect is not prominent.

c. Both the deviance $G^2 = 3.2472$ and Pearson $\chi^2 = 3.2292$ and their respective p-values 0.3551, 0.3576 indicate that $H_0:$ marital happiness is independent of family income cannot be rejected.

d. No. The likelihood ratio test and Wald test both give too large p-values 0.3461, 0.3434 to reject $H_0: \beta = 0$.

e. Average family income means score 2. $P(Y = very) = 1 - \frac{e^{\alpha_2 + \beta * 2}}{1 + e^{\alpha_2 + \beta * 2}} = 0.6132988$.

## 6.11

a. See below:

```{r}
library(MASS)
data <-
array(c(1, 3, 11, 2,
        2, 3, 17, 3,
        0, 1, 8, 5,
        0, 2, 4, 2,
        1, 1, 2, 1,
        0, 3, 5, 1,
        0, 0, 7, 3,
        0, 1, 9, 6),
      dim = c(4, 4, 2),
      dimnames = list(
          JS = c("D", "L", "M", "S"),
          Income = c(3, 10, 20, 35),
          Gender = c("F", "M")))
data.ftable <- ftable(data, row.vars = c("Gender", "Income"))
data.frame.w.freq <- data.frame(data.ftable)
data.frame.w.freq$Income <- as.numeric(data.frame.w.freq$Income)
data.frame.w.freq$Gender <- relevel(data.frame.w.freq$Gender, ref = "M")
m1 <- polr(JS ~ Gender + Income, weights = Freq, data = data.frame.w.freq, Hess = T)
summary(m1)
ctable <- coef(summary(m1))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))
```

The effect of income is statistically prominent. One unit of increase of income would lead to 0.51050833 decrease of the $\log Odds$ of job satisfaction level $Y \le j\ (j = D, L, M)$, which means higher chances of being happier.

b. See below:

```{r}
data.frame.w.freq[data.frame.w.freq$JS == 'L', 'Freq'] <- 
        data.frame.w.freq[data.frame.w.freq$JS == 'D', 'Freq'] +
        data.frame.w.freq[data.frame.w.freq$JS == 'L', 'Freq']
data.frame.w.freq <- data.frame.w.freq[!data.frame.w.freq$JS == 'D', ]
data.frame.w.freq$JS <- as.factor(as.character(data.frame.w.freq$JS))
m2 <- polr(JS ~ Gender + Income, weights = Freq, data = data.frame.w.freq, Hess = T)
summary(m2)
```

Very little is changed, except that there is no intercept for D|L. It shows the cumulative nature of the model. Combining two lower levels won't change the intercepts of levels above them.

c. Dropping gender:

```{r}
data.frame.w.freq <- data.frame(data.ftable)
data.frame.w.freq$Income <- as.numeric(data.frame.w.freq$Income)
m3 <- polr(JS ~ Income, weights = Freq, data = data.frame.w.freq, Hess = T)
summary(m3)
anova(m3, m1)
```

The p-value is too large. Gender does not seem to have a big influence. Can drop.