---
title: "ZhiLiu_HW4"
author: "Zhi Liu"
date: "September 29, 2016"
output: html_document
---

### 5.3

a. That would be comparison between Model 1 and 2.

```{r}
pchisq(173.68 - 170.44, 155 - 152, lower.tail = F)
```

The p-value is large. Therefore the three-factor term can be removed.

b. Compare Model 2 with 3a, 3b and 3c.

```{r}
# 2 and 3a
pchisq(177.34 - 173.68, 158 - 155, lower.tail = F)

# 2 and 3b
pchisq(181.56 - 173.68, 161 - 155, lower.tail = F)

# 2 and 3c
pchisq(173.69 - 173.68, 157 - 155, lower.tail = F)
```

Judging by the large p-value, adding $S * W$ contributes little to Model 3c. Therefore we select it.

c. I'll choose Model 4b $W + C * S$ and take out $C * W$ because of a larger p-value.

d. Compare Model 4b and 5.

```{r}
pchisq(186.61 - 177.61, 166 - 160, lower.tail = F)
```

The p-value is quite small. Should not simplify model.

e. The smallest AIC corresponds to the optimal model according to its own criterion. Therefore Model 5 $C + S + W$.

### 5.19

a. Let $x_i$ represents Department 1 to 5 and $\beta_i (i = 1, ..., 5)$ their effects. Intercept $\alpha$ represents reference level Department 6. We have

$$logit(\pi(Y = 1)) = \alpha + \sum_{i = 1}^5 \beta_i x_i$$

b. Get $\chi^2$ quantile from higher tail.

```{r}
pchisq(21.7, 6, lower.tail = F)
```

THe p-value is very small, meaning the deviance is large. Not a good fit.

c. They signify how far off the observed admittance compared with fitted/expected. Positive means more than expected, and negative means less.

d. The same as the residual for the number of females, because gender is not considered in the fit.

e. The department and gender are highly correlated, and department and admission are highly correlated. Departments like 1 and 2 have a lot of males than females, and these departments tend to admit applicants rather than reject them.