---
title: "ZhiLiu_HW1"
author: "Zhi Liu"
date: "August 28, 2016"
output: html_document
---

### 1.2

a. nominal; b. ordinal; c. ordinal; d. nominal; e. nominal; f. ordinal.

### 1.4

a. When $\pi = 0.50$,

```{r}
dbinom(0, 2, 0.5) # Y = 0
dbinom(1, 2, 0.5) # Y = 1
dbinom(2, 2, 0.5) # Y = 2
```

The mean of the binomial distribution is $n\pi = 2 * 0.5 = 1$. The standard deviation is $\sqrt{n\pi(1-\pi)} = \sqrt{2 * 0.5 * 0.5} = 0.7071068$.

b.
(i) When $\pi = 0.60$,

```{r}
dbinom(0, 2, 0.6) # Y = 0
dbinom(1, 2, 0.6) # Y = 1
dbinom(2, 2, 0.6) # Y = 2
```

(ii) When $\pi = 0.40$,

```{r}
dbinom(0, 2, 0.4) # Y = 0
dbinom(1, 2, 0.4) # Y = 1
dbinom(2, 2, 0.4) # Y = 2
```

c.

$$L(\pi; y=1) = \binom{2}{1} \pi^{2-1} (1-\pi)^1 = 2\pi(1-\pi).$$

```{r}
l <- function(p) {
  2 * p * (1-p)
}
plot(l, c(0.001, 0.999))
```

### 1.6

a. $n_3 = n - n_1 - n_2$.

b. 

\begin{align}
(n_1, n_2, n_3) & = (3, 0, 0) = (0, 3, 0) = (0, 0, 3) \\
                & = (2, 1, 0) = (0, 2, 1) = (1, 0, 2) = (1, 2, 0) = (0, 1, 2) = (2, 0, 1) \\
                & = (1, 1, 1)
\end{align}

c. When $(\pi_1, \pi_2, \pi_3) = (0.25, 0.5, 0.25)$, and $(n_1, n_2, n_3) = (1, 2, 0)$,

```{R}
dmultinom(c(1, 2, 0), prob = c(0.25, 0.5, 0.25))
```

d. $n_1$ alone has a binomial distribution. $n_1$ could be 0, 1, 2. $n_1 \sim Binom(3, 0.25)$.

### 1.8

a. $\hat{p} = 344/1170 = 0.2940171$.

b. Do a proportion test, with $H_0: \pi = 0.5; H_1: \pi > 0$.

```{r}
prop.test(344, 1170, 0.5, conf.level = 0.99, alternative = "greater")
```

Can't reject $H_0$.

Do the same test with $H_0: \pi = 0.5; H_1: \pi < 0$.

```{r}
prop.test(344, 1170, 0.5, conf.level = 0.99, alternative = "less")
```

Can reject $H_0$. Hence the minority would say "yes".

c. This time estimate the two-sided interval with proportion test.

```{r}
prop.test(344, 1170, 0.5, conf.level = 0.99)
```